---
title: "assignment 4"
author: "Marwan Mohamed"
date: "2026-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Excercise 1

## Excercise 2

```{r importing fpp2}
# install has been commented out, please uncomment this and install it if you don't have it
#install.packages("fpp2") 
library(fpp2)

```

1.  What is the frequency of the series? How can you determine this from the data and/or the time-series plot?

    ```{r Part 2 q1}
    # My default way of finding the Frequency 
    frequency(ausbeer)

    # With hindsight, we can also see that the data is split into quarters just by viewing it.
    ausbeer

    # a Subset for easier veiwing and explanation
    ausbeer_subset <- window(ausbeer, start = c(2004, 1), end = c(2007, 1))

    # Plotting the restricted time series
    autoplot(ausbeer_subset) +
      ggtitle("Quarterly Australian Beer Production (2004 - 2007)") +
      xlab("Year") +
      ylab("Megalitres")
    ```

    We can easily view the freqeuncy of the series by using the command frequnecy( ). \
    We can also plainly view the data to see that it is already split into 4 quarters.

    The plot also shows (using a subset for easier viewing) that within every year on the x axis there are 4 distinct points, suggesting that the data is being inputted on quarterly basis.

    Lastly, consider the name of the dataset as introduced in excerice 2, being called "quarterly beer production in Australia" suggesting that the frequency is 4.

    \

2.  Describe the overall behaviour of the series. Is there evidence of a long-term trend?

    ```{r part 2 Q2}
    # Plotting the restricted time series
    autoplot(ausbeer) +
      ggtitle("Quarterly Australian Beer Production") +
      xlab("Year") +
      ylab("Megalitres")
    ```

    The time series plot shows two general phases, a strong positive upward trend from the 1960's till 1973, where the trend disappears and beer production begins to stabilize. There is no evidence of any new trends after\

3.  Is there evidence of seasonality? If so, describe the seasonal pattern.

    ```{r Part 2 Q3 }
    # Plotting the restricted time series... again
    autoplot(ausbeer_subset) +
      ggtitle("Quarterly Australian Beer Production (2004 - 2007)") +
      xlab("Year") +
      ylab("Megalitres")
    ```

    Yes, there is a repeating pattern within each year.\
    The first Quarter is commonly the 2nd most in beer production of that year. production will then drop in Q2 and Q3. After which, production will rise again in Q4 as the best point in beer production of that year. \
    \
    This production increase also aligns with the common holiday season in Australia.\

4.  Does the variability appear constant over time? Explain what you see in the plot and whether a transformation might be appropriate.\
    \
    The variability does not appear to be constant over time. During the start of the upward trend, the seasonal swings was relatively small, around 90 to 100 Megalitires. As overall production increased towards its peak, seasonal swing significantly increased to over 100 consistently and even went above 150 Megalitires on occasion.\
    \
    A transformation would likely be appropriate in this case.\

5.  Are there any observations that you would regard as unusual? Identify them and explain why they attract attention.

    ```{r}

    ```

6.  Plot the series by season (for example, using a seasonal plot). What does this reveal that is less obvious in the ordinary time-series plot?

    ```{r}

    ```

7.  Plot the seasonal subseries. Which quarters have the highest and lowest typical values? Is the seasonal pattern stable over time?

    ```{r}

    ```

8.  Would you detrend or transform the series before modelling it? Give a visual argument for your answer. Do not fit a model.

    ```{r}

    ```
